Sampling Bayesian probabilities given only sampled priors

### 2025-05-25 main.tex
Ready for final reading

\shortauthors{Bernstein et al.}

@@

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eqq}[1]{Equation~(\ref{#1})}
%\tabletypesize{\footnotesize}
\newcommand{\vecc}{\ensuremath{\mathbf{c}}}
\newcommand{\vecq}{\ensuremath{\mathbf{q}}}
\newcommand{\vecs}{\ensuremath{\mathbf{s}}}
\newcommand{\vecn}{\ensuremath{\mathbf{n}}}
\newcommand{\vecu}{\ensuremath{\mathbf{u}}}
\newcommand{\vecU}{\ensuremath{\mathbf{U}}}
\newcommand{\vecv}{\ensuremath{\mathbf{v}}}
\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
\newcommand{\vecP}{\ensuremath{\mathbf{P}}}
\newcommand{\covm}{C}
\newcommand{\matA}{\ensuremath{A}}
\newcommand{\matD}{D}
\newcommand{\matE}{E}
\newcommand{\matF}{F}
\newcommand{\matG}{G}
\newcommand{\matI}{I}
\newcommand{\matT}{T}
\newcommand{\matX}{X}
\newcommand{\matY}{Y}
\newcommand{\matV}{V}
\newcommand{\matLam}{\Lambda}
\newcommand{\proj}{P}  % Projection matrix
\newcommand{\jac}{J}   % Jacobian
\newcommand{\ident}{I}  % identity
\newcommand{\DD}{\Delta_D}
\newcommand{\likeli}{\mathcal{L}}
\newcommand{\trace}{\text{Tr}}
\begin{document}

@@

\include{authors.tex}
% \author[0000-0002-8613-8259]{Gary M. Bernstein}
% \affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
% \email{garyb@upenn.edu}

% \author{William d'Assignies}
% \author{Michael Troxel}
% \author{Alex Alarcon}
% \author{Alex Amon}
% \author{Giulia Giannini}
% \author{Boyan Yin}
% \author{Builders\ldots} %%%%

\begin{abstract}
	\vspace{0.2in}
A typical Bayesian inference on the values of some parameters of
interest $\vecq$ from some data $D$ involves running a Markov Chain (MC) to sample
from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
many cases, the nuisance parameters are high-dimensional, and their
prior $p(\vecn)$ is itself defined only by a set of samples that have
been drawn from some other MC.  Two problems arise: first,
the MC for the posterior will typically require evaluation of
$p(\vecn)$ at arbitrary values of $\vecn,$ \ie\ one needs to provide a
density estimator over the full $\vecn$ space from the provided
samples.  Second, the high dimensionality of $\vecn$ hinders both the
density estimation and the efficiency of the MC for the posterior.  We
describe a solution to this problem: a linear compression of the
$\vecn$ space into a much lower-dimensional space $\vecu$ which
projects away directions in $\vecn$ space that cannot appreciably
alter $\likeli.$ The algorithm for doing so is a slight modification
to principal components analysis, and is less restrictive on
$p(\vecn)$ than other proposed solutions to this issue.
We demonstrate this ``mode projection'' technique using the analysis
of 2-point correlation functions of weak lensing fields and galaxy
density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
$n(z)$ of the galaxies.
\end{abstract}
\reportnum{}

@@

The ability to accomodate non-Gaussian distributions of the
nuisance-parameter space is the principle
advantage of our method over the single-step covariance-inflation
method of \citet{hans}.
But even if the \vecu\ prior is well approximated as multivariate normal,
there are practical advantages of compressing the nuisance 
variables and retaining them in the cosmological Markov chain rather
than using the covariance-inflation method for analytic
marginalization.  One can, for example, examine the posterior
distributions of \vecu\ to see how the data have constrained the prior.  For example, if the cosmological posterior for $\vecu$ is at the edge of the prior, this would potentially indicate an inconsistency between the data and
the prior.

@@

For simplicity, we have demonstrated this method for a case in which
$\vecn$ specifies the $n(z)$ function for a single sample of DES
galaxies.  The method is fully applicable to any set of nuisance
parameters for which we are given a set of samples from $p(\vecn).$
For instance, in the DES Y6 analysis, we use a $\vecn$ that is a
concatenation of the parameters of $n(z)$ for 4 bins of lens source
galaxies.  Since each output sample from the  redshift-estimate
process describes all 4 bins, we know the cross-correlations between
the bins' $n(z)$ behavior, and the compression process described
herein will properly preserve these correlations in the subsequent
cosmological analysis.

\section{Summary}
  We present a linear dimensionality reduction technique that has the
  aim of making it feasible to produce continuous density estimators
  for nuisance parameters that are known only from a set of samples in
  a high-dimensional space $\vecn.$  This is essentially a principle
  components analysis that is adjusted to separate contributions to
  the detectable consequences of $\vecn$ rather than contributions to
  the Euclidean norm of $\vecn.$   This method enables a rigorous
  marginalization over the distribution of $\vecn$ even for
  non-Gaussian distributions, as long as the derivatives of the
  log-likelihood of the data with respect to $\vecn$ are nearly
  constant over the posterior domain.  The mode-projection technique
  is also useful as a method of projecting away irrelevant
  noise fluctuations in $\vecn.$

>>>>

\shortauthors{Bernstein et al.}
\shorttitle{Sampling sampled priors}
\reportnum{DES-2025-0886}
\reportnum{FERMILAB-PUB-25-0362-PPD}

@@

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eqq}[1]{Equation~(\ref{#1})}
%\tabletypesize{\footnotesize}
\newcommand{\vecc}{\ensuremath{\mathbf{c}}}
\newcommand{\vecq}{\ensuremath{\mathbf{q}}}
\newcommand{\vecs}{\ensuremath{\mathbf{s}}}
\newcommand{\vecn}{\ensuremath{\mathbf{n}}}
\newcommand{\vecu}{\ensuremath{\mathbf{u}}}
\newcommand{\vecU}{\ensuremath{\mathbf{U}}}
\newcommand{\vecv}{\ensuremath{\mathbf{v}}}
\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
\newcommand{\vecP}{\ensuremath{\mathbf{P}}}
\newcommand{\covm}{C}
\newcommand{\matA}{\ensuremath{A}}
\newcommand{\matD}{D}
\newcommand{\matE}{E}
\newcommand{\matF}{F}
\newcommand{\matG}{G}
\newcommand{\matI}{I}
\newcommand{\matT}{T}
\newcommand{\matX}{X}
\newcommand{\matY}{Y}
\newcommand{\matV}{V}
\newcommand{\matLam}{\Lambda}
\newcommand{\proj}{P}  % Projection matrix
\newcommand{\jac}{J}   % Jacobian
\newcommand{\ident}{I}  % identity
\newcommand{\DD}{\Delta_D}
\newcommand{\likeli}{\mathcal{L}}
\newcommand{\trace}{\text{Tr}}

\begin{document}

@@

\include{authors.tex}
\correspondingauthor{G.~M.~Bernstein}

\begin{abstract}
	\vspace{0.2in}
A typical Bayesian inference on the values of some parameters of
interest $\vecq$ from some data $D$ involves running a Markov Chain (MC) to sample
from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
many cases, the nuisance parameters are high-dimensional, and their
prior $p(\vecn)$ is itself defined only by a set of samples that have
been drawn from some other MC.  Two problems arise: first,
the MC for the posterior will typically require evaluation of
$p(\vecn)$ at arbitrary values of $\vecn,$ \ie\ one needs to provide a
density estimator over the full $\vecn$ space from the provided
samples.  Second, the high dimensionality of $\vecn$ hinders both the
density estimation and the efficiency of the MC for the posterior.  We
describe a solution to this problem: a linear compression of the
$\vecn$ space into a much lower-dimensional space $\vecu$ which
projects away directions in $\vecn$ space that cannot appreciably
alter $\likeli.$ The algorithm for doing so is a slight modification
to principal components analysis, and is less restrictive on
$p(\vecn)$ than other proposed solutions to this issue.
We demonstrate this ``mode projection'' technique using the analysis
of 2-point correlation functions of weak lensing fields and galaxy
density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
$n(z)$ of the galaxies.
\end{abstract}

@@

The ability to accomodate non-Gaussian distributions of the
nuisance-parameter space is the principle
advantage of our method over the single-step covariance-inflation
method of \citet{hans}.  A further advantage of our
compression$+$density-estimator scheme over covariance inflation is
that nonlinearities in the data model, including presence of
significant $\partial^2\hatc/\partial\vecq\partial\vecn,$ will be
correctly handled in the inference to the extent that the higher
derivatives are confined to the compressed subspace of $\vecn.$ 
But even in the case where linearity and gaussianity in \vecu\ hold,
there are practical advantages of compressing the nuisance 
variables and retaining them in the cosmological Markov chain rather
than using the covariance-inflation method for analytic
marginalization.  One can, for example, examine the posterior
distributions of \vecu\ to see how the data have constrained the prior.  For example, if the cosmological posterior for $\vecu$ is at the edge of the prior, this would potentially indicate an inconsistency between the data and
the prior.

@@

For simplicity, we have demonstrated this method for a case in which
$\vecn$ specifies the $n(z)$ function for a single sample of DES
galaxies.  The method is fully applicable to any set of nuisance
parameters for which we are given a set of samples from $p(\vecn).$
For instance, in the DES Y6 analysis, we use a $\vecn$ that is a
concatenation of the parameters of $n(z)$ for 4 bins of lens source
galaxies.  Since each output sample from the  redshift-estimate
process describes all 4 bins, we know the cross-correlations between
the bins' $n(z)$ behavior, and the compression process described
herein will properly preserve these correlations in the subsequent
cosmological analysis.  In the DES Y6 analysis, we have also extended
the nuisance parameter vector $\vecn$ to include the multiplicative
errors on the shear measurement method, and create a compressed
$\vecu$ that captures correlations between the multiplicative errors
and the $n(z)$ estimates.

\section{Summary}
  We present a linear dimensionality reduction technique that has the
  aim of making it feasible to produce continuous density estimators
  for nuisance parameters that are known only from a set of samples in
  a high-dimensional space $\vecn.$  This is essentially a principle
  components analysis that is adjusted to separate contributions to
  the detectable consequences of $\vecn$ rather than contributions to
  the Euclidean norm of $\vecn.$   This method enables a rigorous
  marginalization over the distribution of $\vecn$ even for
  non-Gaussian distributions, as long as the derivatives of the
  log-likelihood of the data with respect to $\vecn$ are nearly
  constant over the posterior domain.  Indeed the effect of the
  nuisance parameters on the inference will be correctly calculated
  even if these derivatives vary across the posterior domain, as long
  as the statistically relevant values of $\vecn$ do not deviate
  significantly from the linear subspace defined by our compression.
The mode-projection technique
  is also useful as a method of projecting away irrelevant
  noise fluctuations in $\vecn.$
